Modelling Prehistoric Iconographic Compositions. The \proglang{R} package \pkg{decorr}

### 2020-12-10 article_rvTH16.Rnw

%% recommended packages
\usepackage{thumbpdf,lmodern}
\usepackage{float,tikz}
%% another package (only for this demo article)
\usepackage{framed}
%% author pakckage
\usepackage{xcolor,caption, float}

@@

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}
% \setlist[itemize]{noitemsep, topsep=0pt}
<<preliminaries, echo=FALSE, include=FALSE, cache=FALSE >>=
# options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
library("extrafont") #;library("ggplot")
library("igraph")
opts_chunk$set(prompt=TRUE, fig.pos = 'H', results='asis', highlight=FALSE)
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
options(SweaveHooks=list(fig=function() par(mar=c(0, 0, 0, 0))))
# options(replace.assign=TRUE, width=90, prompt="R> ")
opts_knit$set(concordance=TRUE, self.contained=FALSE, background = "#FFFFFF")
# knitr::opts_chunk$set(fig.pos = 'H')
loadfonts(device = "win")
@

@@

\begin{document}
% \SweaveOpts{concordance=TRUE}
% \Sweave2knitr("article_rvTH14_1.Rnw")
% \SweaveOpts{concordance=TRUE}

@@

\smallbreak
<<a_r_t, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = 'H', fig.align="center",fig.asp=0.1, fig.width=10, fig.cap="\\label{fig:fart}Concatenate of 'a', 'r' and 't' graphical units is 'art'">>=
par(mar=c(0.1,0.1,0.1,0.1) )
g <- graph_from_literal(a-+r, r-+t)
plot(g,
     layout = cbind(seq(-1,1,length.out = gorder(g)), 0),
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=80,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier New",
     edge.arrow.size=.5)
@

@@

<<echo=FALSE, message=FALSE, warning=FALSE, fig.pos = 'H', fig.align="center", fig.width=2, fig.height=2, fig.cap="\\label{fig:fratar}Potential spatial relations between graphical units">>=
par(mar=c(0.1,0.1,0.1,0.1) )
ugs <- data.frame(name=c("a","r","t"))
lks <- data.frame(from=c("a", "r", "t"), to=c("r", "t", "a"), type=c("=","=","="))
g <- igraph::graph_from_data_frame(lks, directed=FALSE, vertices= ugs)
plot(g,
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=80,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier New"
)
@

Because of the inherent variability of iconography, and because graphical and spatial proximities between graphical units are generaly not quantified, applying the Saussurian model to any prehistorical graphical content had led to considerable problems:
\begin{itemize}
\setlength\itemsep{.05em}
\item unexplicit groupings of graphical units -- like graphical units grouped into \emp{figures},  \emp{figures} grouped into \emp{patterns}, \emp{patterns} grouped into \emp{motives}, etc. -- with tedious number of groups
\item consistency, proximities and relationships between these groups are often implicit and not quantified
\item studies develop proper descriptive vocabularies, singular relationships of grouping, idosyncratic methods at site-dependend or period-dependend scales 
\end{itemize}

These issues limit drastically the possibility to conduct cross-cultural comparisons and to draw a synthesis of humankind's symbolism at a large scale and over the long-term.  
\\
According to us, prehistorical composition -- like any formal system -- are spatial features (mostly shapes) related one with the other depending on rules of spatial proximities. Then, the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70} should be more operative than the Saussurian '\emph{linearity of the signifier}' principle \citep{Saussure89}. To get these '\emph{near things}', or features proximities, we use concepts coming from the Graph Theory and particularly relative neighborhood graphs (RNG): '\emph{a powerful model of low-level visual process}' \citep{Toussaint80}.  
\\
In this article we present the \proglang{R} package \pkg{decorr}. Its purpose is to formalise a method and offer tools to analyse any graphical content (here: decorations) with methods coming from raph Theory, GIS and binary topological relationships (\emph{birel}) modeling. This package has been grounded on the seminal work of \citet{Alexander08} and its first IT implementation by \citet{Huet18a}.

@@

Graph theory offers a conceptual framework and indices (global at the entire graph scale, local at the vertex scale) to deal with notions of networks, relationships and neighbourhoods. Graphical units (GUs) can be modelled as vertices (nodes) separated one with another by undecorated areas. Their spatial relations can be modelled as edges. The different levels of GUs spatial organisation can be retrieve by a relative neighborhood graphs analysis (Graph Theory) and a spatial (GIS) analysis.\\
% \smallbreak
Nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface. Indeed, for large series of decorated supports, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates. The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis (ie, all the graphical content). The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The polygonisation of the GUs (\code{POLYGONS}), including the border of the stelae, allows to calculated their Voronoi seeds (ie, their centroid) and Voronoi cells (ie, their area of influence). So, the entire decoration image is tiled and condidered as a Voronoi diagram. Centroids of the GUs (\code{POINTS}) are calculated and considered as the graph nodes, their locations (\emph{x} and \emph{y} coordinates, measured in pixels) are relative to each decoration. Exist edges (\code{LINES}) between pairs of GUs when their Voronoi cells share a common border (\emph{birel}: 'touches').\\
The entire graphical content of the decoration is mapped with a neighborhood graphs (RNG) -- ie geometric graph, or planar graph -- of GUs centroids (nodes) and GUs proximity links (edges). So a decoration graph is a one component graph which is the dual graph of the Voronoi diagram (Fig. \ref{fig:gis}).

% \begin{figure}
\begin{figure}[H]
\centering
\includegraphics{gis}
\caption{\label{fig:gis} GIS interface. A) Original decoration of the Late Bronze Age \emp{Cerro Muriano 1} stelae (drawing: \cite{DiazGuardamino10}) opened in new GIS project without any projection system with its extent (\code{xmin, xmax, ymin, ymax} in pixels); B) Polygonsiation of the GUs, calculation of their centroids (nodes), calculation of the Voronoi cells, calculation of the proximity links (edges); C) plot of the relative \emph{x} and \emph{y} coordinates of the GUs centroids; D) plot of the values in the \code{type} of the GUs}
\end{figure}

Whether this theoritical process include image binarization, polygonization and \emph{birel} analysis, a simplier solution will be to create directly centroids on the GUs and to draw the edges manually. Then \emph{x} and \emph{y} coordinates of nodes can be easily retrieved.\\
This model has a minimal of \emph{a priori} definitions. Those definitions only concern the intrinsic variables of GUs (\code{type}, \code{technologie}, \code{color}, \code{size}, etc.) and the types of relations they share. At least one variable, like the \code{type} of the GUs ('anthropomorph', 'horned helmet', etc.), is defined in order to compute composition analysis. Here, we will only consider one property for the nodes, its \code{type}, and the most common types of relations between GUs: \emph{normal}, \emph{attribute} and \emph{overlapping} edges.\\
\\
\textbf{\emph{normal}} edges
\\
By convention, two contemporaneous GUs having a Voronoi cell sharing a border, have a common edge tagged \code{'='}. This proximity is represented with a plain line. The textual notation of such an edge is \code{'-=-'}. For example: \code{personnage -=- bouclier} means that the nodes \code{personnage} and \code{bouclier} have a common border (Fig. \ref{fig:gis_1})  
\smallbreak
\textbf{\emph{attribute}} edges
\\
It occurs frequently that a GU can be divided into a \emph{main unit} (eg, an anthropomorph) and one or various \emph{attribute units} (eg, a helmet, a male sex). Broadly speaking, for further statistical analysis, it is better to use this \emph{attribute} method than to multiply the categories of GUs. To record this information, a new type of edge, tagged with \code{'+'}, is be introduced. This type of edges is directed and, by convention, displayed with a dashed line. Its starts from the \emph{main unit} and ends with the \emph{attribute unit}. The textual notation of such an edge is \code{'-+-'}. For example \code{personnage -+- casque} means that the main node \code{personnage} has the attribute node \code{casque} (Fig. \ref{fig:gis_1}).

\begin{figure}[H] 
\centering
\includegraphics[height=10cm, keepaspectratio]{gis_1}
\caption{\label{fig:gis_1} GIS interface, Cerro Muriano 1 decoration. The GUs \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the GU \code{personnage} (anthropomorph).}
\end{figure}

\smallbreak
\textbf{\emph{overlapping}} edges
\\
Finally, it is quite common that a graphical composition shows superimpositions between different GUs permitting to distinguish different decoration phases, or iconographic layers, for a single support. So, at first, the analyse must be performed on each different phases of decoration separately. This stratigraphical information (\code{A} \emph{over} \code{B}, or \code{B} \emph{under} \code{A}) helps to understand the relative chronology between GUs and must be recorded. A simple way to achieve this is to introduce the new tag \code{'>'} for the for the type of edge. This type of edges is directed. The textual notation of such an edge is \code{'->-'}. For example \code{ecriture ->- bouclier} means that \code{ecriture} crosses \code{bouclier} (ie, \code{ecriture} overlaps \code{bouclier} in the stratigraphical sense) (Fig. \ref{fig:gis_2}).

\begin{figure}[H] 
\centering
\includegraphics[height=10cm, keepaspectratio]{gis_2}
\caption{\label{fig:gis_2} GIS interface, Ibahernando decoration. The GUs \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

These different kind of relationships are resumed with their birel equivalencies in the following table (Tab. \ref{table:tedges})

\begin{table}[H]
% \begin{center}
\centering
\begin{tabular}{|p{.6cm} p{.5cm} p{.6cm} p{2cm} p{2cm} p{5cm}|}
\hline
% node A & type of edge & node B & undirected & birel & stratigraphical meaning \\
GU.1 & edge type & GU.2 & (un)directed & \emph{birel} & description \\
\hline
\code{A} & \code{=} & \code{B} & undirected & $ A \cap B = \emptyset $ & \code{A} and \code{B} are disjoint \code{A} and \code{B} contemporaneous \\
\code{A} & \code{+} & \code{B} & directed & $ A \cap B = A $ & \code{A} and \code{B} are contemporaneous, \code{B} is an attribute of \code{A} \\
\code{A} & \code{>} & \code{B} & directed & $ A \cap B = \exists $ & \code{A} overlaps \code{B}, \code{A} is more recent than \code{B} \\
% \code{1} & = & 2 & undirected & A \cap B = \varnothing & 1 and 2 can be contemporaneous \\
\hline
\end{tabular}
\caption{Synthesis for the different types of relations between GUs}\label{tab1}
% \end{center}
\label{table:tedges}
\end{table}

@@

A training dataset (nodes and edges coordinates, decoration images) is stored in the \code{extdata} folder of the \pkg{decorr} package
\smallbreak
% \begin{itemize}
% \item The \code{imgs} dataframe
% \end{itemize}
\textbf{\code{imgs} dataframe}
\\
The inventory of decorations is stored in the \code{imgs} dataframe (Tab. \ref{tab:timg}). The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (\code{nodes} and \code{edges})
%
<<results='asis', echo=FALSE, warning=FALSE>>=
# library("decorr")
# library("igraph")
# library("magick")
# # par(mar=c(0,0,0,0))
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
print(xtable::xtable(imgs,
                     caption="The studied corpus, the \\code{imgs.tsv} dataframe",
                     label="tab:timg",
                     size=6,
                     align=c(
                       "|p{0.01cm}|",
                       "|p{0.5cm}|",
                       "p{2.5cm}|",
                       "p{2.5cm}|",
                       "p{8.5cm}|")),
      table.placement="H",
      include.rownames=FALSE)
@
%
The \pkg{decorr} package training dataset is composed by five stelaes decorations drawings \citep{DiazGuardamino10} belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- dated to the Late Bronze Age of SW Iberian peninsula \citep{Celestino01}. \\
At first the drawings dataset can be checked by using the \code{imgs} dataframe and the \pkg{magick} package (Fig. \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = 'H', fig.align="center", fig.width=18, fig.height=13, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
# library("magick")
pth <- system.file("extdata", package = "decorr")
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
  lbl.txt <- paste0(imgs[i,"idf"],"\n",
                    imgs[i,"site"],"\n",
                    imgs[i,"decor"],"\n",
                    imgs[i,"img"],"\n",
                    image_info(i1)$width,"*",image_info(i1)$height," px")
  i1 <- image_annotate(i1,lbl.txt,location = "northwest",
                       size = 30, color = "red")
  lims[[length(lims)+1]]<- i1
}
out.img <- image_append(c(image_append(c(lims[[1]],lims[[2]],lims[[3]])),
                          image_append(c(lims[[4]],lims[[5]]))),
                        stack = TRUE)
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{images} dataframe, the first step is to load \code{nodes} and \code{edges} dataframes.
%
<<echo=TRUE>>=
nodes <- read.table(system.file("extdata", "nodes.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
edges <- read.table(system.file("extdata", "edges.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
@
% 
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}
\smallbreak
\textbf{\code{nodes} dataframe}
\\
It contains the required minimum variables for the analysis (Tab. \ref{tab:tnds}).

<<tndss, results='asis', echo=TRUE>>=
caption <- "Nodes (from \\code{nodes.csv} dataframe)}"
print(xtable::xtable(head(nodes),
                     caption = caption,
                     label="tab:tnds"),
      table.placement="H")
@

@@

<<tedss, results='asis', echo=TRUE>>=
caption <- "Edges (from \\code{edges.csv} dataframe)}"
print(xtable::xtable(head(edges),
                     caption=caption,
                     label="tab:teds"),
      table.placement="H")
@

@@

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = 'H', fig.align="center", fig.width=4, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list">>=
lgrph <- list_dec(imgs,nodes,edges)
plot(lgrph[[1]],
     vertex.color = "orange",
     vertex.frame.color="orange",
     vertex.label.color = "black",
     vertex.size = 8,
     vertex.label.cex = 0.6,
     edge.color = "orange",
     vertex.label.family="Helvetica"
)
@

@@

<<nd, results='asis', echo=TRUE, fig.cap="Sixth node of the Torrejon Rubio 1 decoration, the sword">>=
# library(decorr)
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".shp",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(nds.df[6,],
                     # caption="Sixth node (GU 6) of the Torrejon Rubio 1 decoration",
                     label="nd",
                     size=7),
      table.placement="H")
@

@@

<<ed, results='asis', echo=TRUE, fig.cap="First edge of the Torrejon Rubio 1 decoration, the proximity between the sword (GU 6) and the fibula (GU 5)">>=
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(eds.df[1,],
                     # caption="First edge of the Torrejon Rubio 1 decoration, the proximity between the sword (GU 6) and the fibula (GU 5)",
                     label="eds.df",
                     size=7),
      table.placement="H")
@
% \end{figure}

@@

<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = 'H', fig.align="center", fig.height=8, fig.cap="\\label{fig:figs}comparisons between 1 (Cerro Muriano 1) and 4 (Zarza de Montsanchez) decorations">>=
# par(mar=c(0,0,0,0))
g.compar <- list_compar(lgrph,"type")
eds_compar <- plot_compar(g.compar, c(1,4),
                          focus = "edges",
                          doss = system.file("extdata", package = "decorr"))
plot(magick::image_trim(magick::image_read(eds_compar)))
@

>>>>

%% recommended packages
\usepackage{thumbpdf,lmodern}
\usepackage{float,tikz}
%% another package (only for this demo article)
\usepackage{framed}
%% author pakckage
\usepackage{xcolor,caption, float}
\usepackage{array}

@@

%% new custom commands
\newcommand{\class}[1]{`\code{#1}'}
\newcommand{\fct}[1]{\code{#1()}}

%% This modifies which proportion of the page must be occupied by the figure
%% or table to take the full page (with no additional text).
%% Check if the journal does not have any rule about not changing this.
\renewcommand{\floatpagefraction}{.8}

% \setlist[itemize]{noitemsep, topsep=0pt}
<<preliminaries, echo=FALSE, include=FALSE, cache=FALSE >>=
# options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
library("MASS")
library("extrafont") #;library("ggplot")
library("igraph")
opts_chunk$set(prompt=TRUE, fig.pos = '!htbp', results='asis', highlight=FALSE,
               class.source = "bg-success")
options(prompt = "R> ", continue = "+  ", width = 70, useFancyQuotes = FALSE)
options(SweaveHooks=list(fig=function() par(mar=c(0, 0, 0, 0))))
# options(replace.assign=TRUE, width=90, prompt="R> ")
opts_knit$set(concordance=TRUE, self.contained=FALSE, background = "#FFFFFF")
# knitr::opts_chunk$set(fig.pos = '!htbp')
loadfonts(device = "win")
@

@@

\begin{document}
%\SweaveOpts{concordance=TRUE}
% \SweaveOpts{concordance=TRUE}
% \Sweave2knitr("article_rvTH14_1.Rnw")
% \SweaveOpts{concordance=TRUE}

@@

<<a_r_t, echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=1.25, fig.height=0.25, fig.cap="\\label{fig:fart}Concatenate of 'a', 'r' and 't' graphical units is 'art'">>=
par(mar=c(0.1,0.1,0.1,0.1) )
g <- graph_from_literal(a-+r, r-+t)
plot(g,
     layout = cbind(seq(-1,1,length.out = gorder(g)), 0),
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=40,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier",
     edge.arrow.size=0.5,
     xlim=c(-0.94,0.94), 
     ylim=c(-1,-1))
@

@@

<<echo=FALSE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=1.4, fig.height=1.4, fig.cap="\\label{fig:fratar}Potential spatial relations between graphical units">>=
par(mar=c(0,0,0,0))
ugs <- data.frame(name=c("a","r","t"))
lks <- data.frame(from=c("a", "r", "t"), to=c("r", "t", "a"), type=c("=","=","="))
g <- igraph::graph_from_data_frame(lks, directed=FALSE, vertices= ugs)
plot(g,
     vertex.color="white",
     vertex.frame.color="white",
     vertex.size=80,
     vertex.label.cex=1,
     vertex.label.color="black",
     vertex.label.family="Courier",
     xlim=c(-0.65,0.65), 
     ylim=c(-0.65,0.65)
)
@

Because of the inherent variability of iconography, and because graphical and spatial proximities between graphical units are generaly not quantified, applying the Saussurian model to any prehistorical graphical content had led to considerable problems:
\begin{itemize}
\setlength\itemsep{.05em}
\item unexplicit groupings of graphical units -- like graphical units grouped into \emph{figures},  \emph{figures} grouped into \emph{patterns}, \emph{patterns} grouped into \emph{motives}, etc. -- with tedious number of groups
\item consistency, proximities and relationships between these groups are often implicit and not quantified
\item studies develop proper descriptive vocabularies, singular relationships of grouping, idosyncratic methods at site-dependent or period-dependent scales
\end{itemize}

These issues limit drastically the possibility to conduct cross-cultural comparisons and to draw a synthesis of humankind's symbolism at a large scale and over the long-term.

According to us, prehistorical composition -- like any formal system -- are spatial features (mostly shapes) related one with the other depending on rules of spatial proximities. Then, the First Law of Geography: "\emph{everything is related to everything else, but near things are more related than distant things}" \citep{Tobler70} should be more operative than the Saussurian '\emph{linearity of the signifier}' principle \citep{Saussure89}. To get these '\emph{near things}', or features proximities, we use concepts coming from the Graph Theory and particularly relative neighborhood graphs (RNG): '\emph{a powerful model of low-level visual process}' \citep{Toussaint80}.

In this article we present the \proglang{R} package \pkg{decorr}. Its purpose is to formalise a method and offer tools to analyse any graphical content (here: decorations) with methods coming from raph Theory, GIS and binary topological relationships (\emph{birel}) modeling. This package has been grounded on the seminal work of \citet{Alexander08} and its first IT implementation by \citet{Huet18a}.

@@

Graph theory offers a conceptual framework and indices (global at the entire graph scale, local at the vertex scale) to deal with notions of networks, relationships and neighbourhoods. Graphical units (GUs) can be modelled as vertices (nodes) separated one with another by undecorated areas. Their spatial relations can be modelled as edges. The different levels of GUs spatial organisation can be retrieve by a relative neighborhood graphs analysis (Graph Theory) and a spatial (GIS) analysis.

Nodes and edges -- repectively GUs and connexions between GUs -- are created on a GIS interface. Indeed, for large series of decorated supports, GIS offers the most suitable and flexible interface to register all GUs and to get their coordinates. The decoration image is considerated as the basemap of the project and will cover the region of interest of the analysis (ie, all the graphical content). The decoration image can be binarized: GUs are considerated active, the undecorated parts of the support -- the background -- are considerated inactive. The polygonisation of the GUs (\code{POLYGONS}), including the border of the stelae, allows to calculated their Voronoi seeds (ie, their centroid) and Voronoi cells (ie, their area of influence). So, the entire decoration image is tiled and condidered as a Voronoi diagram. Centroids of the GUs (\code{POINTS}) are calculated and considered as the graph nodes, their locations (\emph{x} and \emph{y} coordinates, measured in pixels) are relative to each decoration. Exist edges (\code{LINES}) between pairs of GUs when their Voronoi cells share a common border (\emph{birel}: 'touches').

The entire graphical content of the decoration is mapped with a neighborhood graphs (RNG) -- ie geometric graph, or planar graph -- of GUs centroids (nodes) and GUs proximity links (edges). So a decoration graph is a one component graph which is the dual graph of the Voronoi diagram (Fig. \ref{fig:gis}).

\begin{figure}[!htbp]
\centering
\includegraphics{gis}
\caption{\label{fig:gis} GIS interface. A) Original decoration of the Late Bronze Age \emph{Cerro Muriano 1} stelae (drawing: \cite{DiazGuardamino10}) opened in new GIS project without any projection system with its extent (\code{xmin, xmax, ymin, ymax} in pixels); B) Polygonsiation of the GUs, calculation of their centroids (nodes), calculation of the Voronoi cells, calculation of the proximity links (edges); C) plot of the relative \emph{x} and \emph{y} coordinates of the GUs centroids; D) plot of the values in the \code{type} of the GUs}
\end{figure}

Whether this theoritical process include image binarization, polygonization and \emph{birel} analysis, a simplier solution will be to create directly centroids on the GUs and to draw the edges manually. Then \emph{x} and \emph{y} coordinates of nodes can be easily retrieved.\\
This model has a minimal of \emph{a priori} definitions. Those definitions only concern the intrinsic variables of GUs (\code{type}, \code{technologie}, \code{color}, \code{size}, etc.) and the types of relations they share. At least one variable, like the \code{type} of the GUs ('anthropomorph', 'horned helmet', etc.), is defined in order to compute composition analysis. Here, we will only consider one property for the nodes, its \code{type}, and the most common types of relations between GUs:

\begin{description}
\item[Normal edges]
By convention, two contemporaneous GUs having a Voronoi cell sharing a border, have a common edge tagged \code{'='}. This proximity is represented with a plain line. The textual notation of such an edge is \code{'-=-'}. For example: \code{personnage -=- bouclier} means that the nodes \code{personnage} and \code{bouclier} have a common border (Fig. \ref{fig:gis_1}).

\item[Attribute edges]
It occurs frequently that a GU can be divided into a \emph{main unit} (eg, an anthropomorph) and one or various \emph{attribute units} (eg, a helmet, a male sex). Broadly speaking, for further statistical analysis, it is better to use this \emph{attribute} method than to multiply the categories of GUs. To record this information, a new type of edge, tagged with \code{'+'}, is be introduced. This type of edges is directed and, by convention, displayed with a dashed line. Its starts from the \emph{main unit} and ends with the \emph{attribute unit}. The textual notation of such an edge is \code{'-+-'}. For example \code{personnage -+- casque} means that the main node \code{personnage} has the attribute node \code{casque} (Fig. \ref{fig:gis_1}).

\begin{figure}[!htbp]
\centering
\includegraphics[height=10cm, trim=15mm 20mm 15mm 20mm, clip, keepaspectratio]{gis_1}
\caption{\label{fig:gis_1} GIS interface, Cerro Muriano 1 decoration. The GUs \code{casque} (helmet) and \code{sexe\_masculin} (male sex) are two nodes attributes of the GU \code{personnage} (anthropomorph).}
\end{figure}

\item[Overlapping edges]
Finally, it is quite common that a graphical composition shows superimpositions between different GUs permitting to distinguish different decoration phases, or iconographic layers, for a single support. So, at first, the analyse must be performed on each different phases of decoration separately. This stratigraphical information (\code{A} \emph{over} \code{B}, or \code{B} \emph{under} \code{A}) helps to understand the relative chronology between GUs and must be recorded. A simple way to achieve this is to introduce the new tag \code{'>'} for the for the type of edge. This type of edges is directed. The textual notation of such an edge is \code{'->-'}. For example \code{ecriture ->- bouclier} means that \code{ecriture} crosses \code{bouclier} (ie, \code{ecriture} overlaps \code{bouclier} in the stratigraphical sense) (Fig. \ref{fig:gis_2}).

\end{description}

\begin{figure}[!htbp]
\centering
\includegraphics[height=10cm, trim=30mm 16mm 30mm 13mm, clip, keepaspectratio]{gis_2}
\caption{\label{fig:gis_2} GIS interface, Ibahernando decoration. The GUs \code{ecriture} (writing) overlaps \code{lance} (spear) and the upper part of the \code{bouclier} (shield).}
\end{figure}

These different relationship types are resumed with their birel equivalencies in Table \ref{table:tedges}.

\begin{table}[!htbp]
\centering
\begin{tabular}{cccccp{4.9cm}}
\hline
% node A & type of edge & node B & undirected & birel & stratigraphical meaning \\
GU.1 & Edge type & GU.2 & Edge directness & \emph{Birel} & Description \rule{0pt}{1em} \\[1pt]
\hline \rule{0pt}{1em}%
\code{A} & \code{=} & \code{B} & Undirected & $ A \cap B = \emptyset$ & 
  \code{A} and \code{B} are contemporaneous and disjoint. \\
\code{A} & \code{+} & \code{B} & Directed & $ A \cap B = A $ & 
  \code{A} and \code{B} are contemporaneous. \code{B} is an attribute of \code{A}. \\
\code{A} & \code{>} & \code{B} & Directed & $ A \cap B = \exists $ & 
  \code{A} is more recent than \code{B}.\newline \code{A} overlaps \code{B}. \\
% \code{1} & = & 2 & undirected & A \cap B = \varnothing & 1 and 2 can be contemporaneous \\
\hline
\end{tabular}
\caption{Synthesis for the different types of relations between GUs.}\label{tab1}
\label{table:tedges}
\end{table}

@@

A training dataset (nodes and edges coordinates, decoration images) is stored in the \code{extdata} folder of the \pkg{decorr} package
\smallbreak
% \begin{itemize}
% \item The \code{imgs} dataframe
% \end{itemize}
\textbf{\code{imgs} dataframe}
\\
The inventory of decorations is stored in the \code{imgs} dataframe (Tab. \ref{tab:timg}). The field \code{imgs$idf} is the short name of the decoration, useful during statistical analysis. The primary key of each decoration is the concatenate of \code{imgs$site} and \code{imgs$decor}. These keys will allow joints with the other dataframes (\code{nodes} and \code{edges})
%
<<results='asis', echo=FALSE, warning=FALSE>>=
# library("decorr")
# library("igraph")
# library("magick")
# # par(mar=c(0,0,0,0))
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
print(xtable::xtable(imgs,
    caption="The studied corpus, the \\code{imgs.tsv} dataframe",
    label="tab:timg",
    size=6,
    align=c(
      "p{0.01cm}",
      "c",
      ">{\\raggedright\\arraybackslash\\rule{0pt}{1em}\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
      ">{\\raggedright\\arraybackslash\\addtolength{\\baselineskip}{-1.5pt}}p{2.1cm}",
      "p{8.5cm}")),
    table.placement="!htbp",
    include.rownames=FALSE)
@
%
The \pkg{decorr} package training dataset is composed by five stelaes decorations drawings \citep{DiazGuardamino10} belonging to the so-called 'Warrior stelae' family -- with about 140 stelae -- dated to the Late Bronze Age of SW Iberian peninsula \citep{Celestino01}. \\
At first the drawings dataset can be checked by using the \code{imgs} dataframe and the \pkg{magick} package (Fig. \ref{fig:dataset})
%
<<timg, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=15, fig.height=11.2, fig.cap="\\label{fig:dataset}Decoration images of the training dataset">>=
# library("magick")
pth <- system.file("extdata", package = "decorr")
imgs <- read.table(system.file("extdata", "imgs.tsv", package = "decorr"),
                   sep="\t", stringsAsFactors = FALSE)
lims <- list()
for(i in 1:nrow(imgs)){
  i1 <- image_read(paste0(pth,"\\",imgs[i,"img"]))
  lbl.txt <- paste0(imgs[i,"idf"],"\n",
                    imgs[i,"site"],"\n",
                    imgs[i,"decor"],"\n",
                    imgs[i,"img"],"\n",
                    image_info(i1)$width,"*",image_info(i1)$height," px")
  i1 <- image_annotate(i1,lbl.txt,location = "northwest",
                       size = 28, color = "red")
  lims[[length(lims)+1]]<- i1
}
out.img <- image_append(c(image_append(c(lims[[1]],lims[[2]],lims[[3]])),
                          image_append(c(lims[[4]],lims[[5]]))),
                        stack = TRUE)
par(mar=c(0,0,0,0))
plot(out.img)
@
%
To construct a graph overlapping the decoration images listed in the \code{images} dataframe, the first step is to load \code{nodes} and \code{edges} dataframes.
%
<<echo=TRUE>>=
nodes <- read.table(system.file("extdata", "nodes.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
edges <- read.table(system.file("extdata", "edges.csv", package = "decorr"),
                    sep=";",stringsAsFactors = FALSE)
@
%
% \begin{itemize}
% \item The \code{nodes} dataframe
% \end{itemize}
\smallbreak
\textbf{\code{nodes} dataframe}
\\
It contains the required minimum variables for the analysis (Tab. \ref{tab:tnds}).

<<tndss, results='asis', echo=TRUE>>=
caption <- "Nodes (from \\code{nodes.csv} dataframe)"
print(xtable::xtable(head(nodes),
                     caption = caption,
                     label="tab:tnds"),
      table.placement="!htbp")
@

@@

<<tedss, results='asis', echo=TRUE>>=
caption <- "Edges (from \\code{edges.csv} dataframe)"
print(xtable::xtable(head(edges),
                     caption=caption,
                     label="tab:teds"),
      table.placement="!htbp")
@

@@

<<igraphs, echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=2.5, fig.height=2.5, fig.cap="\\label{fig:igraph}R interface. Plot of the first graph of the list">>=
lgrph <- list_dec(imgs,nodes,edges)
par(mar=c(0,0,0,0))
plot(lgrph[[1]],
     vertex.color = "orange",
     vertex.frame.color="orange",
     vertex.label.color = "black",
     vertex.size = 16,
     vertex.label.cex = 1,
     edge.color = "orange",
     vertex.label.family="Helvetica"
)
@

@@

<<nd, results='asis', echo=TRUE, fig.cap="Sixth node of the Torrejon Rubio 1 decoration, the sword">>=
# library(decorr)
sit <- "Torrejon Rubio" ; dec <- "Torrejon Rubio 1"
nds.df <- read_nds(site = sit, decor = dec, dev = ".shp",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(nds.df[6,],
                     # caption="Sixth node (GU 6) of the Torrejon Rubio 1 decoration",
                     label="nd",
                     size=7),
      table.placement="!htbp")
@

@@

<<ed, results='asis', echo=TRUE, fig.cap="First edge of the Torrejon Rubio 1 decoration, the proximity between the sword (GU 6) and the fibula (GU 5)">>=
eds.df <- read_eds(site = sit, decor = dec, dev = ".tsv",
                   doss = system.file("extdata", package = "decorr"))
print(xtable::xtable(eds.df[1,],
                     # caption="First edge of the Torrejon Rubio 1 decoration, the proximity between the sword (GU 6) and the fibula (GU 5)",
                     label="eds.df",
                     size=7),
      table.placement="!htbp")
@
% \end{figure}

@@

<<echo=TRUE, message=FALSE, warning=FALSE, fig.pos = '!htbp', fig.align="center", fig.width=8, fig.height=4.2, fig.cap="\\label{fig:figs}comparisons between 1 (Cerro Muriano 1) and 4 (Zarza de Montsanchez) decorations">>=
par(mar=c(0,0,0,0))
g.compar <- list_compar(lgrph,"type")
eds_compar <- plot_compar(g.compar, c(1,4),
                          focus = "edges",
                          doss = system.file("extdata", package = "decorr"))
plot(magick::image_trim(magick::image_read(eds_compar)))
@
